NLIZE: A Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models

### 2018-08-01 EMNLP-demo-paper/NLPVis-demo-paper.tex
update figure
minor edits all over
update application

\title{Visual Interrogation of Attention-Based Models: \\ Applications in Language Inference and Machine Comprehension}

@@

\begin{abstract}
Neural networks models have gained unprecedented popularity in natural language processing due to their state-of-the-art performance and the flexible end-to-end training scheme. Despite their advantages, the lack of interpretability hinders the deployment and refinement of the models.
%These models often relying on the attention mechanism, which not only helps improve predicting performance but also provide interpretable representations for users to peak into the opaque models that are usually considered as black boxes.
%
In this work, we present a flexible visualization library for creating customized visual analytic environments, in which the user can investigate and interrogate the relationships among the input, the model internals (i.e., attention), and the output predictions which in turn shed light on the model decision-making processing.

\end{abstract}

@@

@@

>>>>

\title{Visual Interrogation of Attention-Based Models for Natural Language Inference and Machine Comprehension}

@@

\begin{abstract}
Neural networks models have gained unprecedented popularity in natural language processing due to their state-of-the-art performance and the flexible end-to-end training scheme. Despite their advantages, the lack of interpretability hinders the deployment and refinement of the models. In this work, we present a flexible visualization library for creating customized visual analytic environments, in which the user can investigate and interrogate the relationships among the input, the model internals (i.e., attention), and the output predictions which in turn shed light on the model decision-making processing.
\end{abstract}

@@

\section*{Acknowledgments}
This work was performed under the auspices of the U.S. Department of Energy by Lawrence Livermore National Laboratory under Contract DE-AC52-07NA27344.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2018-08-01 paper/application.tex
fit into the limit
update application

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{att2pred}
 \vspace{-6mm}
 \caption{
Editing the original attention (a) to correctly align the word "heap" with "pile" as shown in (b) (these two words are highlighted in orange).
The change of attention leads to the change of prediction from neutral to the class boundary between \emph{neutral} and \emph{entitlement}.
%
}
\label{fig:att2pred}
\end{figure}

% perturb input, attention
\begin{figure*}[t]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{pipelineUpdate}
  \vspace{-6mm}
 \caption{
Experiment with all configurations for the label reassignment optimization. As shown in (d), the update to the attention stage seems to have significantly less impact on the prediction result compared to the classifier or encoder stage of the model.
 }
  \vspace{-4mm}
\label{fig:pipelineUpdate}
\end{figure*}

@@

Interestingly, all configurations except one are concentrated around the full \emph{neutral} prediction. Referring back to the pipeline visualization, we observe that the only configuration that failed to produce the correct label is the one for which we allow only updating of the attention stage of the model.
%
%In other words, the updating of the attention stage seems to have significantly less impact on the prediction result compared to the classifier or encoder stage of the model.
%Similar observations can be obtained for most examples we have experimented with.
%\shusen{
%The domain experts comment that this observation is interesting and worth further investigation from the perspective of NLP research.
%The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, while the composed semantics participate in classification layer. From this analysis, we can extrapolate that adjusting the word alignment affects only the composed semantics but has no effect on word semantics, which often plays a significant role in a good model (as supported by recent works in word embeddings, e.g., ELMo~\cite{salant2017contextualized}). Without proper semantics (in both encoder and classifier), only adjusting alignment is not sufficient to yield good model. 
%
The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, whereas the composed semantics participate in the classification layer. From this analysis, we can extrapolate that the ultimate decision relies on the semantics (i.e., encodings and composed encodings in the classifier). The encoder and classifier layer can swiftly adjust weights for word/composed semantics to correct prediction. However, the attention layer, which only works on how individual words are aligned, affects the prediction less significantly. Recent word embeddings works (e.g., ELMo~\cite{salant2017contextualized}) also support such an observation that better word embeddings can substantially benefit a model. However, the result by no means implies attention is not useful, because it serves as a way to compose word semantics.
%To interpret the result, we need to carry out a case by case analysis, as the separation of the stages depends on the individual implementation.
%In our case, the result indicates that the tunable parameters of the attention stage (the transformation before the alignment operation, see details in Appendix A) has less impact on the final prediction compared to the other two stages.
%However, this result does not necessarily imply the limitation of attention mechanism as the core alignment computation does not have tunable parameters that can be modified by the proposed optimization.
%}
%The attention layer is less sensitive compared to encoding layer and classification layer.
%When examined other sentences pairs, such an observation remains.
%This gives domain expert some hint about the significance of the role of each pipeline stage play in this model.
% which based on the natural extension of the margin-infused relaxed algorithm (MIRA) to the neural network generalize the concept of maligin combined effort of the  proposed technique

%
%However, to understand how the model should
\begin{figure*}[t]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{depTreeExample}
  \vspace{-6mm}
 \caption{
The dependency tree provides valuable information that can help fix the prediction error.
In (a), the model mistakenly aligns the word green, which leads to an incorrect prediction.
After examining the dependency tree (highlighted by pink squares), we can see the two \textbf{greens} are attached to different words.
In (b), by editing the attention and forcing the alignment of the two \textbf{greens} to be zero, the prediction label is corrected to \emph{neutral}.
 }
\label{fig:depTreeExample}
 \vspace{-4mm}
\end{figure*}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=0.9\linewidth]{att2pred}
 \vspace{-4mm}
 \caption{
Editing the original attention (a) to correctly align the word "heap" with "pile" as shown in (b) (these two words are highlighted in orange).
The change of attention leads to the change of prediction from neutral to the class boundary between \emph{neutral} and \emph{entitlement}.
%
}
\label{fig:att2pred}
\vspace{-3mm}
\end{figure}

% perturb input, attention
\begin{figure*}[t]
\centering
\vspace{-3mm}
 \includegraphics[width=0.95\linewidth]{pipelineUpdate}
  \vspace{-2mm}
 \caption{
Experiment with all configurations for the label reassignment optimization. As shown in (d), the update to the attention stage seems to have significantly less impact on the prediction result compared to the classifier or encoder stage of the model.
 }
  \vspace{-4mm}
\label{fig:pipelineUpdate}
\end{figure*}

@@

Interestingly, all configurations except one are concentrated around the full \emph{neutral} prediction. Referring back to the pipeline visualization, we observe that the only configuration that failed to produce the correct label is the one for which we allow only updating of the attention stage of the model.
%
%In other words, the updating of the attention stage seems to have significantly less impact on the prediction result compared to the classifier or encoder stage of the model.
%Similar observations can be obtained for most examples we have experimented with.
%\shusen{
%The domain experts comment that this observation is interesting and worth further investigation from the perspective of NLP research.
%The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, while the composed semantics participate in classification layer. From this analysis, we can extrapolate that adjusting the word alignment affects only the composed semantics but has no effect on word semantics, which often plays a significant role in a good model (as supported by recent works in word embeddings, e.g., ELMo~\cite{salant2017contextualized}). Without proper semantics (in both encoder and classifier), only adjusting alignment is not sufficient to yield good model. 
%
The domain experts find this observation very interesting and suggest a preliminary interpretation. The reason, as they believe, is that the attention layer functions differently from the encoder and the classifier in terms of word semantics. The attention layer in this model is more specialized in composing word semantics rather than creating new semantics. When it comes to fixing a wrong prediction, attention layer affects the prediction less significantly while the encoder and the classifier can swiftly adapt to the correct label. Thus from this analysis, we can extrapolate that the ultimate prediction relies on the semantics (i.e., encodings and composed encodings in the classifier). Recent word embeddings works (e.g., ELMo~\cite{salant2017contextualized}) also support such an observation that better word embeddings can substantially benefit a model. However, the result by no means implies attention is not useful, because it serves as a way to compose word semantics.
%To interpret the result, we need to carry out a case by case analysis, as the separation of the stages depends on the individual implementation.
%In our case, the result indicates that the tunable parameters of the attention stage (the transformation before the alignment operation, see details in Appendix A) has less impact on the final prediction compared to the other two stages.
%However, this result does not necessarily imply the limitation of attention mechanism as the core alignment computation does not have tunable parameters that can be modified by the proposed optimization.
%}
%The attention layer is less sensitive compared to encoding layer and classification layer.
%When examined other sentences pairs, such an observation remains.
%This gives domain expert some hint about the significance of the role of each pipeline stage play in this model.
% which based on the natural extension of the margin-infused relaxed algorithm (MIRA) to the neural network generalize the concept of maligin combined effort of the  proposed technique

%
%However, to understand how the model should
\begin{figure*}[t]
\centering
\vspace{-2mm}
 \includegraphics[width=0.98\linewidth]{DepTreeExample}
  \vspace{-3mm}
 \caption{
The dependency tree provides valuable information that can help fix the prediction error.
In (a), the model mistakenly aligns the word green, which leads to an incorrect prediction.
After examining the dependency tree (highlighted by pink squares), we can see the two \textbf{greens} are attached to different words.
In (b), by editing the attention and forcing the alignment of the two \textbf{greens} to be zero, the prediction label is corrected to \emph{neutral}.
 }
\label{fig:depTreeExample}
 \vspace{-4mm}
\end{figure*}
